Follow-up Samples

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
75% completed results modification

# Main CRC bacteria differences between Adenoma and Carcinoma
main_crc_bugs_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)

@@

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was no difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  This change can be visualized using an NMDS [Figure 1c & 1d]. When all follow up samples were compared to each other there was no significant overall difference between them (P-value = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

***Outcome of Model Training***
The range of the AUC for model training ranged from a minimum of `r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)` to a maximum of `r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)` with the middle of all `r length(rownames(model_train_summary))` runs having an AUC of `r format(model_test_AUC_summary["middle", "ROC_cv"], 
digits = 3)`.  Interestingly, the worst AUC model from training performed the best on it's respective 80/20 split test data [Figure 3].  In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, middle training model test set AUC = `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, worse training model test set AUC = `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`).  That is to say it had the smallest change in AUC in comparison to the minimum and maximum AUC trained models.  The middle model was close to the full training data AUC `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`. There was no significant difference between the AUC of the best and middle training models (P-value = `r format(roc_pvalues["best", "middle"], digits = 3)`).  There was also no difference in the middle model versus worse (P-value = `r format(roc_pvalues["middle", "worse"], digits = 3)`) or full data model (P-value = `r format(roc_pvalues["middle", "full"], digits = 3)`).  The two comparisons with a significant difference were between the worse training model and the best training model (P-value = `r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)`) and the worse training model and full training model (P-value = `r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)`).

@@

***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S3].  A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.  In general, `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was decreased from initial to follow up [Figure 5].  The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].

***Differences in Adenoma and Carcinoma in Previously Associated Cancer Bacteria***
First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].  The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].

>>>>

# Main CRC bacteria differences between Adenoma and Carcinoma
main_crc_bugs_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
crc_bug_info <- read.csv("../results/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)

@@

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery are visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was also no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
We next examined whether there were and differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4]. Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

***Full and Reduced Model Results***
Since differences were observed between initial and follow up samples and only a small number of individuals were positive for previously associated CRC bacteria we next investigated if we could create a model that could adequately classify and adjust lesion probability based on the bacterial community and FIT.  The lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` to `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` versus the initial follow up model which had and AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 20 repeated 10-fold cross validations.  Interestingly, identification of the most important variables and reducing the variables considered to only these increased the AUC in the lesion model (`r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) and initial and follow up model (`r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).

The test set AUC range for the full and reduced lesion model were similar to that reported for the training set AUC ranges and the ROC curve ranges overlap each other [Figure 3a].  The ROC curve for the final lesion model used falls within the range of both the full and reduced lesion model [Figure 3a].

@@
